---
title: "personalactivity"
author: "Peri"
date: "September 10, 2016"
output: html_document
---
##Set up the environment and load the required libraries
```{r}
library(knitr)
opts_chunk$set(echo = TRUE, results = 'hold')
library(data.table)
library(ggplot2)

```

#1.Read the data
```{r}
mydata <- read.csv('activity.csv', header = TRUE, sep = ",",
                  colClasses=c("numeric", "character", "numeric"))

```
##process the data
```{r}
mydata$date <- as.Date(mydata$date, format = "%Y-%m-%d")
mydata$interval <- as.factor(mydata$interval)
```
##let us see how the data looks
```{r}
str(mydata)
```
#2.Histogram of the total number of steps taken each day
##find the total number of steps in order to plot them
```{r}
stepsperday <- aggregate(steps ~ date, mydata, sum)
colnames(stepsperday) <- c("date","steps")
```
#plot the data
```{r}
ggplot(stepsperday, aes(x = steps)) + 
       geom_histogram(fill = "green", binwidth = 1000) + 
        labs(title="Histogram of Steps Taken per Day", 
             x = "Number of Steps per Day", y = "Number of times in a day") + theme_bw() 
```


#3. Mean and median number of steps taken each day
```{r}
meansteps   <- mean(stepsperday$steps, na.rm=TRUE)
mediansteps <- median(stepsperday$steps, na.rm=TRUE)
```
#4.Time series plot of the average number of steps taken
##to find the aggregate of steps in interval of 5 minutes 
```{r}
stepsperinterval <- aggregate(mydata$steps, 
                                by = list(interval = mydata$interval),
                                FUN=mean, na.rm=TRUE)
stepsperinterval$interval <- 
        as.integer(levels(stepsperinterval$interval)[stepsperinterval$interval])
colnames(stepsperinterval) <- c("interval", "steps")
```
##plot the time series found 
```{r}
ggplot(stepsperinterval, aes(x=interval, y=steps)) +   
        geom_line(color="blue", size=1) +  
        labs(title="Average steps taken", x="Interval", y="Number of steps") +  
        theme_bw()
```


#5.The 5-minute interval that, on average, contains the maximum number of steps
```{r}
maxstepsinterval <- stepsperinterval[which.max(  
        stepsperinterval$steps),]
```

#6.Code to describe and show a strategy for imputing missing data
##Let us find the total number of missing values first
```{r}
missingvals <- sum(is.na(mydata$steps))
```
##come up with the strategy to fill the missing data
#first we create a function which takes the mydata and steps_per_interval as arguments. This function goes through each value per index and when it finds a missing value, we replace with the mean value
```{r}
nafill <- function(data, pervalue) {
        naindex <- which(is.na(data$steps))
        nareplace <- unlist(lapply(naindex, FUN=function(idx){
                interval = data[idx,]$interval
                pervalue[pervalue$interval == interval,]$steps
        }))
        fillsteps <- data$steps
        fillsteps[naindex] <- nareplace
        fillsteps
}

mydatafill <- data.frame(  
        steps = nafill(mydata, stepsperinterval),  
        date = mydata$date,  
        interval = mydata$interval)
str(mydatafill)
```
##let us check if there are any missing values after replacing with mean
```{r}
sum(is.na(mydatafill$steps))
```
#7.Histogram of the total number of steps taken each day after missing values are imputed
```{r}
fill_steps_per_day <- aggregate(steps ~ date, mydatafill, sum)
colnames(fill_steps_per_day) <- c("date","steps")

ggplot(fill_steps_per_day, aes(x = steps)) + 
       geom_histogram(fill = "blue", binwidth = 1000) + 
        labs(title="Histogram of Steps Taken per Day", 
             x = "Number of Steps per Day", y = "Number of times in a day") + theme_bw() 
```


#8.Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
##to find the difference let us find the mean and median of steps after filling the missing values
```{r}
steps_mean_fill   <- mean(fill_steps_per_day$steps, na.rm=TRUE)
steps_median_fill <- median(fill_steps_per_day$steps, na.rm=TRUE)
```
##let us compare the patterns between weekdays and weekends
##In order to do so, to give the days of the week
```{r}
weekdayssteps <- function(data) {
    weekdayssteps <- aggregate(data$steps, by=list(interval = data$interval),
                          FUN=mean, na.rm=T)
    # convert to integers for plotting
    weekdayssteps$interval <- 
            as.integer(levels(weekdayssteps$interval)[weekdayssteps$interval])
    colnames(weekdayssteps) <- c("interval", "steps")
    weekdayssteps
}

```
##to subset the data into weekdays and weekends for the comparison and find the average steps per interval
```{r}
data_by_weekdays <- function(data) {
    data$weekday <- 
            as.factor(weekdays(data$date))
    weekenddata <- subset(data, weekday %in% c("Saturday","Sunday"))
    weekdaydata <- subset(data, !weekday %in% c("Saturday","Sunday"))

    weekendsteps <- weekdayssteps(weekenddata)
    weekdaysteps <- weekdayssteps(weekdaydata)

    weekendsteps$dayofweek <- rep("weekend", nrow(weekendsteps))
    weekdaysteps$dayofweek <- rep("weekday", nrow(weekdaysteps))

    data_by_weekdays <- rbind(weekendsteps, weekdaysteps)
    data_by_weekdays$dayofweek <- as.factor(data_by_weekdays$dayofweek)
    data_by_weekdays
}

dataonweekdays <- data_by_weekdays(mydatafill)
```
##plot the data
```{r}
ggplot(dataonweekdays, aes(x=interval, y=steps)) + 
        geom_line(color="violet") + 
        facet_wrap(~ dayofweek, nrow=2, ncol=1) +
        labs(x="Interval", y="Number of steps") +
        theme_bw()
```

